---
title: "LiDAR CHM Visualization"
author: "Conor McMahon"
date: "11/1/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(raster)
library(rgdal)
library(here)
library(rasterKernelEstimates)

```

First, load the two LiDAR canopy height models and display them for visual comparison, after running a filter to remove powerlines:

First dataset information:
- first flight on October 18, 2015 the second on October 19, 2015, and a short final mission was flown on October 20, 2015
- 1 km wide, 125 km long, about 136 km^3 total area
- 10 to 45 pts/m^2
Second dataset information:
- flights from 05/27/2018 - 07/22/2018
- about 7423 km^2
- 5.92 pts/m^3

```{r load_filter_chms, echo=FALSE, warning=FALSE}

pre_drought_chm <- raster(here::here("data","chm_mosaic_2015.tif"))
post_drought_chm <- raster(here::here("data","chm_mosaic_2018.tif"))

# Filter to remove power lines
#    get PERCENTILEth percentile of heights among neighbors to each pixel
#    assign NA to pixels which have height more than HEIGHT_THRESHOLD above the PERCENTILEth percentile in neighbors
#    neighbors are all points within KERNEL_HALFWIDTH cells (including along diagonals) except target point
filterPowerLines <- function(input_raster, kernel_halfwidth, percentile, height_threshold)
{
  # Weights matrix is for 24 nearest neighbors
  kernel_width <- (kernel_halfwidth*2+1)
  weights <- matrix(1,nrow=kernel_width, ncol=kernel_width)
  weights[kernel_halfwidth+1,kernel_halfwidth+1] <- 0
  run.time <- proc.time()
  kernel_percentiles <- rasterLocalQuantiles(input_raster,weights,q=percentile)
  
  powerline_mask <- (input_raster - kernel_percentiles) > height_threshold
  
  return(mask(input_raster, powerline_mask, maskvalue=1, updatevalue=NA))
}

# Here, using 70th percentile, kernels 25 pixels on a side (25 m), and a height threshold of 10
pre_drought_chm <- filterPowerLines(pre_drought_chm, 12, 70, 10)
writeRaster(pre_drought_chm, here::here("outputs","chm_mosaic_2015_filtered.tif"), overwrite=TRUE)
post_drought_chm <- filterPowerLines(post_drought_chm, 12, 70, 10)
writeRaster(post_drought_chm, here::here("outputs","chm_mosaic_2018_filtered.tif"), overwrite=TRUE)

plot(pre_drought_chm, zlim=c(0,30))
plot(post_drought_chm, zlim=c(0,30))

```

Next, aggregate to five meters and mask both images to the same extent, then compare changes:

```{r aggregate_mask_change_detection, echo=FALSE, warning=FALSE}

source(here::here("aggregate_custom.R"))

# Aggregate up to 5m resolution
crs(pre_drought_chm) <- crs("+init=epsg:32611")
pre_drought_chm_5m <- aggregate(pre_drought_chm, fact=5, fun=mean)
post_drought_chm_5m <- aggregate_custom(post_drought_chm, pre_drought_chm_5m)

pre_drought_mask <- pre_drought_chm_5m != 0
post_drought_mask <- post_drought_chm_5m != 0

# plot(pre_drought_chm_5m)
# plot(post_drought_chm_5m)
# plot(pre_drought_mask)
# plot(post_drought_mask)

# Get Change in CHM from 2015 to 2018
drought_change <- (post_drought_chm_5m - pre_drought_chm_5m) * pre_drought_mask * post_drought_mask
drought_change[drought_change == 0] <- NA
plot(drought_change, zlim=c(-10,10))

# Crop scenes to match same extent
drought_change <- crop(drought_change, pre_drought_chm_5m)
pre_drought_chm_5m <- crop(pre_drought_chm_5m, drought_change)

pre_drought_chm_df <- as.data.frame(pre_drought_chm_5m, xy=TRUE)
names(pre_drought_chm_df) <- c("x", "y", "chm_init")
drought_change_df <- as.data.frame(drought_change, xy=TRUE)
names(drought_change_df) <- c("x", "y","chm_change")
drought_change_df$chm_init <- pre_drought_chm_df$chm_init

ggplot(drought_change_df) + 
  geom_histogram(aes(x=chm_change))

```

Compare change between sites



```{r site_level_change, echo=FALSE, warning=FALSE}

woodland <- readOGR(here::here("boundaries","east_grove_woodland.shp"))
channel <- readOGR(here::here("boundaries","channel_SCR.shp"))

channel_change <- mask(drought_change, channel)
woodland_change <- mask(drought_change, woodland)

writeRaster(channel_change, here::here("outputs","channel_change"), overwrite=TRUE)
writeRaster(woodland_change, here::here("outputs","woodland_change"), overwrite=TRUE)

channel_change_df <- as.data.frame(channel_change, xy=TRUE) %>%
  drop_na() %>%
  mutate(scene = rep("channel", n()))
woodland_change_df <- as.data.frame(woodland_change, xy=TRUE) %>%
  drop_na() %>%
  mutate(scene = rep("woodland", n()))
change_df <- rbind(channel_change_df, woodland_change_df)

ggplot(change_df) + 
  geom_histogram(aes(x=layer, group=scene, col=scene), bins=100) + 
  scale_x_continuous(limits=c(-30,15)) + 
  scale_y_log10()


```


Interested in comapring histograms across time...


```{r histogram_correlations, warning=FALSE}

# Load the two datasets
histogram_2015 <- stack(here::here("outputs","histogram_2015_mosaic.tif"))
histogram_2018 <- stack(here::here("outputs","histogram_2018_mosaic.tif"))
# Crop to same extent
histogram_2015 <- crop(histogram_2015, histogram_2018)
histogram_2018 <- crop(histogram_2018, histogram_2015)
# Mask to remove values which are empty in one raster
mask_2015 <- 1-is.na(histogram_2015[[1]])
mask_2018 <- 1-is.na(histogram_2018[[1]])
mask_overall <- mask_2015*mask_2018
histogram_2015 <- histogram_2015*mask_overall
histogram_2018 <- histogram_2018*mask_overall

# Combine histogram values to one image (first two bands relate to min/max of histogram range...)
histograms_combined <- stack(histogram_2015[[3:22]], histogram_2018[[3:22]])

# Sourced code for this part from obrl_soil answer at https://gis.stackexchange.com/questions/279535/pixel-correlations-from-two-raster-datasets-in-r

# Function to generate correlation between 20 initial and 20 following values for a given pixel
corvec <- function(vec = NULL) {
  cor(
    # first LiDAR histogram (2015)
    x      = vec[1:(length(vec)/2)],
    # second LiDAR histogram (2018)
    y      = vec[((length(vec)/2) + 1):length(vec)],
    use    = 'complete.obs',
    method = 'spearman'
  )
}

# Run the above function for each cell in the raster
histogram_correlation <- calc(
  histograms_combined,
  fun = function(x) {
    # skip all areas where at least one band in the cell is NA in at least one of the images
    if (sum(is.na(x)) > 0) {
      NA_real_
    } else {
      corvec(vec = x)
    }
  }
)

plot(histogram_correlation, zlim=c(0,1))

writeRaster(histogram_correlation, here::here("outputs","histogram_correlation.tif"), overwrite=TRUE)

woodland_histogram_correlation <- mask(histogram_correlation, woodland)
channel_histogram_correlation <- mask(histogram_correlation, channel)
writeRaster(woodland_histogram_correlation, here::here("outputs","woodland_histogram_correlation"), overwrite=TRUE)
writeRaster(channel_histogram_correlation, here::here("outputs","channel_histogram_correlation"), overwrite=TRUE)

```

