---
title: "Site Specific LiDAR Data"
author: "Conor McMahon"
date: "2/23/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(raster)

```

We're going to extract LiDAR statistics around nest points and survey points for both 2015 and 2018.

First, we'll define some functions to get the height histogram in a local region around a point, and also a histogram of the CHM around each point. 

```{r lidar_stat_functions, include=FALSE, warn=FALSE}
{
  
  getNeighborhood <- function(input_image, point, buffer_radius)
  {
    point <- spTransform(point, crs(input_image))
    masked_data <- raster::extract(input_image, point, buffer=buffer_radius)
  }
}

```

Load LiDAR data - both CHM rasters (including maximum height in each 1 m pixel) and histograms with all return values (including number of returns at various heights within each 10 m pixel).

```{r load_rasters, include=FALSE}

# Load LiDAR raster data
histogram_2015 <- stack(here::here("outputs","histogram_2015_mosaic.tif"))
histogram_2018 <- stack(here::here("outputs","histogram_2018_mosaic.tif"))
chm_2015 <- raster(here::here("outputs","chm_mosaic_2015_filtered.tif"))
chm_2018 <- raster(here::here("outputs","chm_mosaic_2018_filtered.tif"))

```


Analyze survey points, using 50 m radii around each point. First we'll look only at maximum returns within each 1m pixel window:

```{r chm_analysis, include=FALSE}

# Load and apply CRS to survey points (Lat/Lon values)
survey_points <- read_csv(here::here("birds","Point_count_station_list.csv"))
coordinates(survey_points) <- ~Longitude+Latitude
crs(survey_points) <- crs("+init=epsg:4326")

# Get all CHM values within 50 m radius of each survey point
#   In 2015:
survey_max_returns_2015 <- getNeighborhood(chm_2015, survey_points, buffer_radius = 50)
#   In 2018:
survey_max_returns_2018 <- getNeighborhood(chm_2018, survey_points, buffer_radius = 50)

# Bin to a histogram of maximum return height within each 1 m pixel in 50 m radius around a survey points
binLiDARReturns <- function(height_vector, min_height, max_height, bins, year)
{
  # Upper end of each bin
  bin_breaks <- (1:bins)*(max_height-min_height)/bins
  # Lower end of each bin
  bin_starts <- (1:bins-1)*(max_height-min_height)/bins
  # Generate histogram data frame
  histogram_df <- data.frame(height = height_vector, 
                             bin_index = .bincode(height_vector, bin_starts)) %>%
    mutate(bin_start = bin_starts[bin_index],
           bin_end = bin_breaks[bin_index]) %>%
    group_by(bin_index) %>% 
    summarize(count = n(),
              bin_index = median(bin_index),
              bin_start = median(bin_start),
              bin_end = median(bin_end)) %>% 
    mutate(year = rep(year, n()))
  # Add normalized frequency
  histogram_df$frequency <- histogram_df$count / sum(histogram_df$count)
  return(histogram_df)
}

# Apply CHM (max height) binning function for each survey point
#   In 2015:
chm_histograms_2015 <- lapply(survey_max_returns_2015, binLiDARReturns, min_height=0, max_height=30, bins=15, year=2015) 
#   In 2018:
chm_histograms_2018 <- lapply(survey_max_returns_2018, binLiDARReturns, min_height=0, max_height=30, bins=15, year=2018)
# The above are lists of data frames... add survey point labels to each data frame:
for(ind in 1:length(chm_histograms_2015))
{
  chm_histograms_2015[[ind]]$site <- rep(survey_points[ind,]$`Station #`, nrow(chm_histograms_2015[[ind]]))
  chm_histograms_2018[[ind]]$site <- rep(survey_points[ind,]$`Station #`, nrow(chm_histograms_2018[[ind]]))
}
# Combine list of dataframes to one dataframe
chm_histograms_all <- rbind(bind_rows(chm_histograms_2015), bind_rows(chm_histograms_2018))

# Generate plot of CHM histograms by site and year
ggplot(chm_histograms_all) + 
    theme_bw() +
  theme(strip.background = element_rect(fill = 'gray80')) + 
    geom_line(aes(x=bin_start, y=frequency, group=year, col=as.factor(year))) + 
    facet_wrap(~site) + 
    scale_x_continuous(limits=c(0,20)) + 
    ggtitle("Histograms of Maximum LiDAR Returns by Year") + xlab("Height (m)") +
    ylab("Frequency") + 
    guides(col=guide_legend(title="Year"))

```

Now we want to repeat the above, but considering ALL returns, not just the highest ones:

```{r all_returns_analysis, include=FALSE}

# Get matrices where rows rows are pixels in the histogram image, columns are height bins, and cells are frequencies
#    In 2015:
hist_neighborhood_2015 <- getNeighborhood(histogram_2015, survey_points, buffer_radius = 50)
#    In 2018
hist_neighborhood_2018 <- getNeighborhood(histogram_2018, survey_points, buffer_radius = 50)
# Convert the above to a list of height frequency averages across all neighbor pixels in image:
collapseNeighborhood <- function(neighborhood)
{
  # Drop first two columns (which include min/max values in histogram), sum across pixels (columns)
  num_bins <- ncol(neighborhood) - 2
  frequency_sums <- as.numeric(colSums(neighborhood))[3:(num_bins+2)]
  normalized_frequency <- frequency_sums / (sum(frequency_sums))
  # Set up bin to height conversion
  min_value <- median(neighborhood[,1])
  max_value <- median(neighborhood[,2])
  bin_breaks = (1:num_bins)*(max_value-min_value)/num_bins + min_value
  bin_starts = (0:(num_bins-1))*(max_value-min_value)/num_bins + min_value
  bin_indices = 1:num_bins
  return ( data.frame(bin_start = bin_starts,
                      bin_break = bin_breaks, 
                      bin_index = bin_indices,
                      frequency = normalized_frequency) )
}
return_frequencies_2015 <- lapply(hist_neighborhood_2015, collapseNeighborhood)
return_frequencies_2018 <- lapply(hist_neighborhood_2018, collapseNeighborhood)
# Add site information
for(ind in 1:length(return_frequencies_2015))
{
  return_frequencies_2015[[ind]]$site <- rep(survey_points[ind,]$`Station #`, nrow(return_frequencies_2015[[ind]]))
  return_frequencies_2015[[ind]]$year <- rep(2015, nrow(return_frequencies_2015[[ind]]))
  return_frequencies_2018[[ind]]$site <- rep(survey_points[ind,]$`Station #`, nrow(return_frequencies_2018[[ind]]))
  return_frequencies_2018[[ind]]$year <- rep(2018, nrow(return_frequencies_2015[[ind]]))
}
# Combine list of dataframes to one dataframe
total_histograms_all <- rbind(bind_rows(return_frequencies_2015), bind_rows(return_frequencies_2018))

# Replace 0 frequencies with NA to clarify graphs
total_histograms_all[total_histograms_all$frequency == 0,]$frequency <- NA

# Generate plot of all return values by site and year
ggplot(total_histograms_all) + 
    theme_bw() +
  theme(strip.background = element_rect(fill = 'gray80')) + 
    geom_line(aes(x=bin_start, y=frequency, group=year, col=as.factor(year))) + 
    facet_wrap(~site) + 
    scale_x_continuous(limits=c(0,20)) + 
    ggtitle("Histograms of All LiDAR Returns by Year") + xlab("Height (m)") +
    ylab("Frequency") + 
    guides(col=guide_legend(title="Year"))

```


Now lets do some comparisons to bird data:

```{r bird_comparison, include=FALSE}

birds <- rbind(read_csv(here::here("birds","SP_abundance_breeding_2015_2021.csv")),
               read_csv(here::here("birds","SP_abundance_nonbreeding_2015_2021.csv"))) %>%
  janitor::clean_names()
# Bird sites use a different naming convention than survey points - correct here (e.g. PC-01 is SP 1)
site_name_remapping <- c(paste("SP ",1:9,sep=""), paste("SP",10:15,sep=""))
names(site_name_remapping) <- c(paste("PC-0",1:9,sep=""),paste("PC-",10:15,sep=""))
birds$site <- as.character(site_name_remapping[birds$site_code])

# Reformat histograms to be wide:
chm_histograms_wide <- chm_histograms_all %>% 
  dplyr::select(bin_index,frequency,year,site) %>%          # drop extra bin information
  drop_na() %>%                                             
  pivot_wider(names_from=bin_index, values_from=frequency)  # switch to wide format - bins as columns
# Re-order columns so bins are in sensible order
chm_histograms_wide <- chm_histograms_wide %>% 
  dplyr::select(c(1,2,2+(order(as.numeric(colnames(chm_histograms_wide)[3:ncol(chm_histograms_wide)]))))) %>%
  replace(is.na(.), 0)

total_histograms_wide <- total_histograms_all %>% 
  dplyr::select(bin_index,frequency,year,site) %>%          # drop extra bin information
  drop_na() %>%                                             
  pivot_wider(names_from=bin_index, values_from=frequency)  # switch to wide format - bins as columns
# Re-order columns so bins are in sensible order
total_histograms_wide <- total_histograms_wide %>% 
  dplyr::select(c(1,2,2+(order(as.numeric(colnames(chm_histograms_wide)[3:ncol(total_histograms_wide)]))))) %>%
  replace(is.na(.), 0)


yewa_data <- birds %>% 
  filter(year %in% c(2015, 2018),
         species == "YEWA")

bird_lidar_data <- merge(chm_histograms_wide, yewa_data, by=c("year","site"))
names(bird_lidar_data)  <- c("site","year",
                  paste("bin_",names(bird_lidar_data[3:16]),sep=""),
                  "site_code", "species", "total_birds", "relative_abund_index")

```



