---
title: "Site Specific LiDAR Data"
author: "Conor McMahon"
date: "2/23/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(raster)
library(rgdal)

```

We're going to extract LiDAR statistics around survey points for both 2015 and 2018.

First, we'll define some functions to get the height histogram in a local region around a point, and also a histogram of the CHM around each point. 

```{r lidar_stat_functions, warning=FALSE}
{
  
  getNeighborhood <- function(input_image, point, buffer_radius)
  {
    # reproject POINT to raster CRS
    point <- spTransform(point, crs(input_image))
    # extract data in BUFFER_RADIUS distance around POINT
    masked_data <- raster::extract(input_image, point, buffer=buffer_radius)
  }
}

```

Load LiDAR data - both CHM rasters (including maximum height in each 1 m pixel) and histograms with all return values (including number of returns at various heights within each 10 m pixel).

```{r load_rasters, warning=FALSE}

# Load LiDAR raster data
histogram_2015 <- stack(here::here("outputs","histogram_2015_mosaic.tif"))
histogram_2018 <- stack(here::here("outputs","histogram_2018_mosaic.tif"))
chm_2015 <- raster(here::here("outputs","chm_mosaic_2015_filtered.tif"))
crs(chm_2015) <- crs("+init=epsg:32611")
chm_2018 <- raster(here::here("outputs","chm_mosaic_2018_filtered.tif"))

```


Analyze survey points, using 50 m radii around each point. First we'll look only at maximum returns within each 1m pixel window:

```{r chm_analysis, warning=FALSE}

# Load and apply initial CRS to survey points (Lat/Lon values)
survey_points <- read_csv(here::here("birds","Point_count_station_list.csv"))
coordinates(survey_points) <- ~Longitude+Latitude
crs(survey_points) <- crs("+init=epsg:4326")

# Get all CHM values within a fixed radius of each survey point
#   For HR points, birds were detected within a 100 m radius, so that's what we use here
#   For SP points, birds were detected within a 50 m radius, so again that's what we'll use here. 
#   In 2015:
survey_max_returns_2015_hr <- getNeighborhood(chm_2015, survey_points[survey_points$Project == "HR",], buffer_radius = 100)
survey_max_returns_2015_sp <- getNeighborhood(chm_2015, survey_points[survey_points$Project == "SP",], buffer_radius = 50)
survey_max_returns_2015 <- c(survey_max_returns_2015_hr, survey_max_returns_2015_sp)
#   In 2018:
survey_max_returns_2018_hr <- getNeighborhood(chm_2018, survey_points[survey_points$Project == "HR",], buffer_radius = 100)
survey_max_returns_2018_sp <- getNeighborhood(chm_2018, survey_points[survey_points$Project == "SP",], buffer_radius = 50)
survey_max_returns_2018 <- c(survey_max_returns_2018_hr, survey_max_returns_2018_sp)

# Function to bin to a histogram of maximum return height within each 1 m pixel in fixed radius around a survey points
binLiDARReturns <- function(height_vector, min_height, max_height, bins, year)
{
  # Upper end of each bin
  bin_breaks <- (1:bins)*(max_height-min_height)/bins
  # Lower end of each bin
  bin_starts <- (1:bins-1)*(max_height-min_height)/bins
  # Generate histogram data frame
  histogram_df <- data.frame(height = height_vector, 
                             bin_index = .bincode(height_vector, bin_starts)) %>%
    mutate(bin_start = bin_starts[bin_index],
           bin_end = bin_breaks[bin_index]) %>%
    group_by(bin_index) %>% 
    summarize(count = n(),
              bin_index = median(bin_index),
              bin_start = median(bin_start),
              bin_end = median(bin_end)) %>% 
    mutate(year = rep(year, n()))
  # Add normalized frequency
  histogram_df$frequency <- histogram_df$count / sum(histogram_df$count)
  return(histogram_df)
}

# Apply CHM (max height) binning function for each survey point
#   In 2015:
chm_histograms_2015 <- lapply(survey_max_returns_2015, binLiDARReturns, min_height=0, max_height=30, bins=15, year=2015) 
#   In 2018:
chm_histograms_2018 <- lapply(survey_max_returns_2018, binLiDARReturns, min_height=0, max_height=30, bins=15, year=2018)
# The above are lists of data frames... add survey point labels to each data frame:
for(ind in 1:length(chm_histograms_2015))
{
  chm_histograms_2015[[ind]]$site <- rep(survey_points[ind,]$`Station #`, nrow(chm_histograms_2015[[ind]]))
  chm_histograms_2018[[ind]]$site <- rep(survey_points[ind,]$`Station #`, nrow(chm_histograms_2018[[ind]]))
}
# Combine list of dataframes to one dataframe
chm_histograms_all <- rbind(bind_rows(chm_histograms_2015), bind_rows(chm_histograms_2018))

# Generate plot of CHM histograms by site and year
ggplot(chm_histograms_all) + 
    theme_bw() +
  theme(strip.background = element_rect(fill = 'gray80')) + 
    geom_line(aes(x=bin_start, y=frequency, group=year, col=as.factor(year))) + 
    facet_wrap(~site) + 
    scale_x_continuous(limits=c(0,20)) + 
    ggtitle("Histograms of Maximum LiDAR Returns by Year") + xlab("Height (m)") +
    ylab("Frequency") + 
    guides(col=guide_legend(title="Year"))

```

Now we want to repeat the above, but considering ALL returns, not just the highest ones:

```{r all_returns_analysis, warning=FALSE}

# Get all CHM values within a fixed radius of each survey point
#   For HR points, birds were detected within a 100 m radius, so that's what we use here
#   For SP points, birds were detected within a 50 m radius, so again that's what we'll use here. 
#   In 2015:
hist_neighborhood_2015_hr <- getNeighborhood(histogram_2015, survey_points[survey_points$Project == "HR",], buffer_radius = 100)
hist_neighborhood_2015_sp <- getNeighborhood(histogram_2015, survey_points[survey_points$Project == "SP",], buffer_radius = 50)
hist_neighborhood_2015 <- c(hist_neighborhood_2015_hr, hist_neighborhood_2015_sp)
#   In 2018:
hist_neighborhood_2018_hr <- getNeighborhood(histogram_2018, survey_points[survey_points$Project == "HR",], buffer_radius = 100)
hist_neighborhood_2018_sp <- getNeighborhood(histogram_2018, survey_points[survey_points$Project == "SP",], buffer_radius = 50)
hist_neighborhood_2018 <- c(hist_neighborhood_2018_hr, hist_neighborhood_2018_sp)
# Convert the above to a list of height frequency averages across all neighbor pixels in image:
collapseNeighborhood <- function(neighborhood)
{
  # Drop first two columns (which include min/max values in histogram), sum across pixels (columns)
  num_bins <- ncol(neighborhood) - 2
  frequency_sums <- as.numeric(colSums(neighborhood))[3:(num_bins+2)]
  normalized_frequency <- frequency_sums / (sum(frequency_sums))
  # Set up bin to height conversion
  min_value <- median(neighborhood[,1])
  max_value <- median(neighborhood[,2])
  bin_breaks = (1:num_bins)*(max_value-min_value)/num_bins + min_value
  bin_starts = (0:(num_bins-1))*(max_value-min_value)/num_bins + min_value
  bin_indices = 1:num_bins
  return ( data.frame(bin_start = bin_starts,
                      bin_break = bin_breaks, 
                      bin_index = bin_indices,
                      frequency = normalized_frequency) )
}
return_frequencies_2015 <- lapply(hist_neighborhood_2015, collapseNeighborhood)
return_frequencies_2018 <- lapply(hist_neighborhood_2018, collapseNeighborhood)
# Add site information
for(ind in 1:length(return_frequencies_2015))
{
  return_frequencies_2015[[ind]]$site <- rep(survey_points[ind,]$`Station #`, nrow(return_frequencies_2015[[ind]]))
  return_frequencies_2015[[ind]]$year <- rep(2015, nrow(return_frequencies_2015[[ind]]))
  return_frequencies_2018[[ind]]$site <- rep(survey_points[ind,]$`Station #`, nrow(return_frequencies_2018[[ind]]))
  return_frequencies_2018[[ind]]$year <- rep(2018, nrow(return_frequencies_2015[[ind]]))
}
# Combine list of dataframes to one dataframe
total_histograms_all <- rbind(bind_rows(return_frequencies_2015), bind_rows(return_frequencies_2018))

# Replace 0 frequencies with NA to clarify graphs
total_histograms_all[total_histograms_all$frequency == 0,]$frequency <- NA

# Generate plot of all return values by site and year
ggplot(total_histograms_all) + 
    theme_bw() +
  theme(strip.background = element_rect(fill = 'gray80')) + 
    geom_line(aes(x=bin_start, y=frequency, group=year, col=as.factor(year))) + 
    facet_wrap(~site) + 
    scale_x_continuous(limits=c(0,20)) + 
    ggtitle("Histograms of All LiDAR Returns by Year") + xlab("Height (m)") +
    ylab("Frequency") + 
    guides(col=guide_legend(title="Year"))

# Now, resample all the bins to a few 'ecologically meaningful' bins representing different strata used by various guilds of birds:
#   bin 1 = 0 to 2 m (understory)
#   bin 2 = 2 to 4 m (midstory)
#   bin 3 = 4 to 8 m (lower canopy)
#   bin 4 = 8+ m     (upper canopy)
total_histograms_all_eco_bins <- total_histograms_all %>% 
    pivot_wider(names_from=bin_start, names_prefix = "bin_", values_from=frequency) %>%
    group_by(site, year) %>%
    summarize(bin_1 = sum(bin_0,bin_1, na.rm=TRUE),
              bin_2 = sum(bin_2,bin_3, na.rm=TRUE),
              bin_3 = sum(bin_4,bin_5,bin_6,bin_7, na.rm=TRUE),
              bin_4 = sum(bin_8,bin_9,bin_10,bin_11,bin_12,bin_13,bin_14,bin_15,bin_16,bin_17,bin_18,bin_19,bin_20,bin_21,bin_22,bin_23,bin_24,bin_25,bin_26,bin_27,bin_28,bin_29, na.rm=TRUE)) %>%
    pivot_longer(3:6, names_to="bin_name", values_to="frequency") %>%
    mutate(bin = as.numeric(substr(bin_name,5,100)))

```


Now lets do some comparisons to bird data:

```{r bird_comparison, warning=FALSE}

birds <- rbind(read_csv(here::here("birds","SP_abundance_breeding_2015_2021.csv")) ) %>%#,
   #            read_csv(here::here("birds","SP_abundance_nonbreeding_2015_2021.csv"))) %>%
  janitor::clean_names()
# Bird sites use a different naming convention than survey points - correct here (e.g. PC-01 is SP 1)
site_name_remapping <- c(paste("SP ",1:9,sep=""), paste("SP",10:15,sep=""))
names(site_name_remapping) <- c(paste("PC-0",1:9,sep=""),paste("PC-",10:15,sep=""))
birds$site <- as.character(site_name_remapping[birds$site_code])

# Reformat histograms to be wide:
total_histograms_wide <- total_histograms_all %>% 
  dplyr::select(bin_index,frequency,year,site) %>%          # drop extra bin information
  drop_na() %>%                                             
  pivot_wider(names_from=bin_index, values_from=frequency)  # switch to wide format - bins as columns
# Re-order columns so bins are in sensible order
total_histograms_wide <- total_histograms_wide %>% 
  dplyr::select(c(1,2,2+(order(as.numeric(colnames(total_histograms_wide)[3:ncol(total_histograms_wide)]))))) %>%
  replace(is.na(.), 0)

total_histograms_wide <- total_histograms_all %>% 
  dplyr::select(bin_index,frequency,year,site) %>%          # drop extra bin information
  drop_na() %>%                                             
  pivot_wider(names_from=bin_index, values_from=frequency)  # switch to wide format - bins as columns
# Re-order columns so bins are in sensible order
total_histograms_wide <- total_histograms_wide %>% 
  dplyr::select(c(1,2,2+(order(as.numeric(colnames(chm_histograms_wide)[3:ncol(total_histograms_wide)]))))) %>%
  replace(is.na(.), 0)


# Lets pick some target birds based on how many records there are for the species (can't represent species that are never detected)
bird_data_summary <- birds %>% 
  group_by(species) %>%
  summarize(total_count = sum(total_birds),
            avg_abund_index = mean(relative_abund_index),
            presence_count = n())
view(bird_data_summary)

# Now we'll print a bunch of species-specific models...

# Function to build a linear regression between LiDAR values and species abundance
buildBirdModel <- function(target_species)
{
  print(paste("Working on linear model for ", target_species, sep=""))
  
  # Get species-level data
  species_data <- birds %>% 
    filter(year %in% c(2015, 2018),  # filter to 2015 and 2018 (LiDAR overflight years)
           species == target_species)        # filter to target species

  # Combine bird and LiDAR datasets
  bird_lidar_data <- merge(total_histograms_wide, species_data, by=c("site","year"), all.x=TRUE) %>% 
    filter(substr(site,1,1)=="S") %>% 
    replace(is.na(.), 0)
  # Make bin names non-numeric so lm() isn't angry at us
  names(bird_lidar_data)  <- c("site","year",
                    paste("bin_",names(bird_lidar_data[3:(ncol(bird_lidar_data)-5)]),sep=""),
                    "site_code", "species", "total_birds", "relative_abund_index")
  # make sub-bins because we don't have much training data
  # rn, binning like this:
  #   bin 1 -> 0    to 1.5 m
  #   bin 2 -> 1.5  to 7.5 m
  #   bin 3 -> 7.5  to 13.5 m
  #   bin 4 -> 13.5 to 20.0 m
  bird_lidar_data$quad_1 <- bird_lidar_data$bin_1 + bird_lidar_data$bin_2
  bird_lidar_data$quad_2 <- bird_lidar_data$bin_3 + bird_lidar_data$bin_4 
  bird_lidar_data$quad_3 <- bird_lidar_data$bin_5 + bird_lidar_data$bin_6 + bird_lidar_data$bin_7 + bird_lidar_data$bin_8
  bird_lidar_data$quad_4 <- rowSums(bird_lidar_data[,11:(ncol(bird_lidar_data)-8)])
  # Build and summarize a linear model
  bird_model <- lm(data=bird_lidar_data, relative_abund_index~quad_1+quad_2+quad_3+quad_4-1)
  print(summary(bird_model))
  bird_lidar_data$relative_abund_index_predicted <- predict(bird_model,bird_lidar_data)
  # plot prediction results results
  ggplot(data = bird_lidar_data) + 
    geom_point(aes(x=relative_abund_index, y=relative_abund_index_predicted, col=as.factor(year))) + 
    ggtitle(paste("Model Efficacy for ", target_species, sep="")) + 
    ylab("Predicted Relative Abundance Index") + 
    xlab("Actual Relative Abundance Index")
}


buildBirdModel("PSFL")
buildBirdModel("YEWA")
buildBirdModel("WIWA")
buildBirdModel("COYE")
buildBirdModel("YBCH")
buildBirdModel("LBVI")
buildBirdModel("WAVI")
buildBirdModel("PUFI")
buildBirdModel("SOSP")
buildBirdModel("SWTH")
buildBirdModel("HAWO")
buildBirdModel("TRES")

```

Test something similar, looking at distributions of vegetation around nest locations? 


```{r chm_analysis, warning=FALSE}

# Load and apply initial CRS to survey points (Lat/Lon values)
survey_points <- read_csv(here::here("birds","HRNA_Taylor_nests_2015_2021.csv"))
coordinates(survey_points) <- ~Longitude+Latitude
crs(survey_points) <- crs("+init=epsg:4326")
survey_points_sf <- st_as_sf(survey_points)

# Load study site polygons
study_regions <- readOGR(here::here("boundaries","study_sites_feb25.shp"))
woodland <- st_as_sf(study_regions[study_regions$name == "Woodland",])
channel <- st_as_sf(study_regions[study_regions$name == "Channel",])
restoration <- st_as_sf(study_regions[study_regions$name == "Restoration",])
# Add site information to nest dataframe
survey_points_regions <- sf::st_join(survey_points_sf, st_transform(woodland, st_crs(survey_points_sf)))
survey_points_regions <- survey_points_regions[!is.na(survey_points_regions$name),]
temp <- sf::st_join(survey_points_sf, st_transform(channel, st_crs(survey_points_sf)))
survey_points_regions <- rbind(survey_points_regions, 
                               temp[!is.na(temp$name),])
temp <- sf::st_join(survey_points_sf, st_transform(restoration, st_crs(survey_points_sf)))
survey_points_regions <- rbind(survey_points_regions, 
                               temp[!is.na(temp$name),])

# Add new study site labels to original data frame
survey_points_regions <- merge(survey_points, 
                               as.data.frame(survey_points_regions[,c("name",
                                                                      "Nest Id labels for projects")]), 
                               by="Nest Id labels for projects") 
# Drop geometry information, rename 'name' to 'site'
survey_points_regions <- survey_points_regions[,1:7]
survey_points_regions$site <- survey_points_regions$name
survey_points_regions <- survey_points_regions[,-(ncol(survey_points_regions)-1)]

# Get all CHM values within 50 m radius of each survey point
#   In 2015:
survey_max_returns_2015 <- getNeighborhood(chm_2015, survey_points, buffer_radius = 50)
#   In 2018:
survey_max_returns_2018 <- getNeighborhood(chm_2018, survey_points, buffer_radius = 50)

# Function to bin to a histogram of maximum return height within each 1 m pixel in 50 m radius around a survey points
binLiDARReturns <- function(height_vector, min_height, max_height, bins, year)
{
  # Upper end of each bin
  bin_breaks <- (1:bins)*(max_height-min_height)/bins
  # Lower end of each bin
  bin_starts <- (1:bins-1)*(max_height-min_height)/bins
  # Generate histogram data frame
  histogram_df <- data.frame(height = height_vector, 
                             bin_index = .bincode(height_vector, bin_starts)) %>%
    mutate(bin_start = bin_starts[bin_index],
           bin_end = bin_breaks[bin_index]) %>%
    group_by(bin_index) %>% 
    summarize(count = n(),
              bin_index = median(bin_index),
              bin_start = median(bin_start),
              bin_end = median(bin_end)) %>% 
    mutate(year = rep(year, n()))
  # Add normalized frequency
  histogram_df$frequency <- histogram_df$count / sum(histogram_df$count)
  return(histogram_df)
}

# Apply CHM (max height) binning function for each survey point
#   In 2015:
chm_histograms_2015 <- lapply(survey_max_returns_2015, binLiDARReturns, min_height=0, max_height=30, bins=15, year=2015) 
#   In 2018:
chm_histograms_2018 <- lapply(survey_max_returns_2018, binLiDARReturns, min_height=0, max_height=30, bins=15, year=2018)
# The above are lists of data frames... add survey point labels to each data frame:
for(ind in 1:length(chm_histograms_2015))
{
  chm_histograms_2015[[ind]]$nest <- rep(survey_points[ind,]$`Nest Id labels for projects`, nrow(chm_histograms_2015[[ind]]))
  chm_histograms_2018[[ind]]$nest <- rep(survey_points[ind,]$`Nest Id labels for projects`, nrow(chm_histograms_2018[[ind]]))
  
  chm_histograms_2015[[ind]]$nest_year <- rep(survey_points[ind,]$`Year`, nrow(chm_histograms_2015[[ind]]))
  chm_histograms_2018[[ind]]$nest_year <- rep(survey_points[ind,]$`Year`, nrow(chm_histograms_2018[[ind]]))
  
  chm_histograms_2015[[ind]]$site <- rep(survey_points_regions[ind,]$`site`, nrow(chm_histograms_2015[[ind]]))
  chm_histograms_2018[[ind]]$site <- rep(survey_points_regions[ind,]$`site`, nrow(chm_histograms_2018[[ind]]))
}
# Combine list of dataframes to one dataframe
chm_histograms_all <- rbind(bind_rows(chm_histograms_2015), bind_rows(chm_histograms_2018))

# Generate plot of CHM histograms by nest and year
ggplot(chm_histograms_all[chm_histograms_all$site == "Channel",]) + 
    theme_bw() +
    theme(strip.background = element_rect(fill = 'gray80')) + 
    geom_line(aes(x=bin_start, y=frequency, group=year, col=as.factor(year))) + 
    facet_wrap(~nest_year+nest) + 
    scale_x_continuous(limits=c(0,20)) + 
    ggtitle("Histograms of Maximum LiDAR Returns by Year") + xlab("Height (m)") +
    ylab("Frequency") + 
    guides(col=guide_legend(title="Year"))

```
